---
title : "Day 1 // Hands-on 2 // Explore SingleCellExperiment"
author: "Frederick J Tan"
date  : "1 October 2021"
output: 
  html_notebook:
    toc: true
    toc_float: true
---

# Load packages

```{r message=FALSE}
library( "Seurat" )
library( "scater" )
library( "scran" )

library( "tidyverse" )
```

# Load data

## Seurat -> SingleCellExperiment

```{r}
sce <- readRDS( "/home/idies/workspace/practical_genomics/day1/retina.rds" )
sce <- as.SingleCellExperiment( sce )
```

## Check data

```{r}
sce
```

# 1. assay()

## Introduce

- How to access the expression data?

```{r}
help( assay )
```

```{r}
m_assay <- assay( sce, "logcounts" )
```

- What class of object is this?

```{r}
class( m_assay )
```

- How large is it?

```{r}
dim( m_assay )
```

- How to subset first 10 rows, first 3 columns?

```{r}
m_assay[ 1:10, 1:3 ]
```

## Subset RHO

```{r}
x_rho <- m_assay[ "RHO", ]
```

```{r}
class( x_rho )
```

```{r}
length( x_rho )
```

```{r}
hist( x_rho )
```

```{r}
summary( x_rho )
```

## Subset RHO, VIM

```{r}
goi <- c( "RHO", "VIM" )
x_rhovim <- m_assay[ goi, ]
```

```{r}
class( x_rhovim )
```

```{r}
dim( x_rhovim )
```

```{r}
hist( x_rhovim )
```

## ggplot RHO, VIM

- `t()`, `data.frame()`, `rownames_to_column()`, `pivot_longer()`

```{r}
df_rhovim <- t( x_rhovim )
df_rhovim <- data.frame( df_rhovim )
df_rhovim <- rownames_to_column( df_rhovim )
df_rhovim <- pivot_longer( df_rhovim, cols=c( "RHO", "VIM" ) )
df_rhovim
```

```{r}
ggplot( df_rhovim ) +
  geom_histogram( aes( value ) ) +
  facet_wrap( vars(name) )
```

# 2. colData()

## Introduce

- How to access the cell metadata?

```{r}
help( colData )
```

```{r}
meta <- colData( sce )
```

- What class of object is this?

```{r}
class( meta )
```

- What is DFrame?

```{r}
meta
data.frame( meta )
```

- How large is it?

```{r}
dim( meta )
```

- What categories exist?

```{r}
colnames( meta )
```

## Subset rows

- `colData()`, `data.frame()`, `rownames_to_column()`

```{r}
df_meta <- colData( sce ) %>%
  data.frame() %>%
  rownames_to_column()
df_meta
```

```{r}
dim( df_meta )
```

```{r}
df_meta[ 1:10, 1:3 ]
```

## Subset libraryID

```{r}
table( df_meta$libraryID )
```

```{r}
df_fovea2 <- df_meta %>%
  filter( libraryID=="fovea_donor_2" )
df_fovea2
```

```{r}
dim( df_fovea2 )
```

```{r}
table( df_fovea2$cell_type )
```

## ggplot gene by cell_type

- `full_join()`, `filter()`

```{r}
df_rhovim_fovea2 <- df_rhovim %>%
  full_join( df_meta, by="rowname" ) %>%
  filter( libraryID=="fovea_donor_2" )
df_rhovim_fovea2
```

- `geom_violin()`, `facet_wrap()`, `theme()`

```{r}
ggplot( df_rhovim_fovea2 ) +
  geom_violin( aes( cell_type, value ) ) +
  facet_wrap( vars(name) ) +
  theme( axis.text.x=element_text( angle=90 ) )
```

# 3. reducedDim()

## Introduce

- How to access the cluster visualization?

```{r}
help( reducedDim )
```

- `data.frame()`, `rownames_to_column()`

```{r}
df_umap <- reducedDim( sce, "UMAP" ) %>%
  data.frame() %>%
  rownames_to_column()
df_umap
```

## ggplot cell_type

```{r}
df_umap_meta <- full_join( df_umap, df_meta, by="rowname" )
df_umap_meta
```

```{r}
ggplot( df_umap_meta ) +
  geom_point( aes( UMAP_1, UMAP_2, color=cell_type ) )
```

## ggplot RHO

```{r}
df_umap_rho <- full_join( df_umap, df_rhovim, by="rowname" ) %>%
  filter( name=="RHO" )
df_umap_rho
```

```{r}
ggplot( df_umap_rho ) +
  geom_point( aes( UMAP_1, UMAP_2, color=value ) )
```

# 4. findMarkers()

## Introduce

- How to access differentially expressed genes?

```{r}
help( findMarkers )
```

- Preview of findMarkers() syntax

```{r}
colLabels( sce ) <- colData( sce )$cell_type
markers <- findMarkers( sce )
```

- What class of object is this?

```{r}
class( markers )
```

- How large is it?

```{r}
length( markers )
```

- Build up to `data.frame( markers[[1]] )`

```{r}
markers
```

## hist summary.logFC

```{r}
markers[1]
```

```{r}
df_markers_amacrine <- markers[[1]] %>%
  data.frame()
df_markers_amacrine
```

```{r}
hist( df_markers_amacrine$summary.logFC )
```

## plotUMAP markers

```{r}
df_markers_amacrine %>%
  filter( summary.logFC > 0 )
```

```{r}
help( plotUMAP )
```

```{r fig.width=10}
plotUMAP( sce, colour_by="MEG3", text_by="cell_type" )
```

# Exercise 2

1. Recreate plotUMAP() using ggplot2
    - Plot another gene, cell metadata
1. Recreate plotExpression() using ggplot2
    - Plot another gene by cell metadata
1. Adjust ggplot2 appearance
    - Add titles, customize x/y labels, theme
1. Explore findMarkers() results
    - Use plotUMAP(), plotExpression()

# Document software

```{r}
sessionInfo()
```


