---
title : "Day 1 // Exercise 1 // Retina Dataset"
author: ""
date  : "04 October 2021"
output: 
  html_notebook:
    toc: true
    toc_float: true
---

# Summary

- Published as [Voigt et al., 2019](https://pubmed.gov/31075224/)

# Load packages

```{r message=FALSE}
library( "Seurat" )
library( "scater" )
```

# Load data

```{r}
sce <- readRDS( "/home/idies/workspace/practical_genomics/day1/retina.rds" )
sce <- as.SingleCellExperiment( sce )
```

# Check data

```{r}
sce
```

# Explore data

## plotUMAP -- metadata

```{r}
plotUMAP( sce, colour_by="cell_type" )
```

## plotUMAP -- gene

```{r}
plotUMAP( sce, colour_by="RHO" )
```

## plotExpression -- distribution

```{r}
plotExpression( sce, "RHO", "cell_type" ) +
  theme( axis.text.x=element_text( angle=90 ) )
```

# Task 1: Plot metadata

- Here are the available metadata categories

```{r}
colnames( colData(sce) )
```

- Modify plotUMAP() to colour_by a different category

```{r}
plotUMAP( sce, colour_by="" )
```

- Repeat to find one or two patterns you find interesting

```{r}
plotUMAP()
```

# Task 2: Plot gene

- Use this chunk to check the spelling of your gene of interest

```{r}
goi <- "RHO"
table( goi %in% rownames(sce) )
```

- Modify plotUMAP() to plot your gene of interest

```{r}
plotUMAP( sce, colour_by="RHO" )
```

- Repeat to find one or two patterns you find interesting

```{r}

```

# Task 3: Plot distributions

- Modify plotExpression() to compare expression across different categories

```{r}
plotExpression( sce, "RHO", "libraryID" ) +
  theme( axis.text.x=element_text( angle=90 ) )
```

- Replace `libraryID` with your category to tabulate how many cells are in your category

```{r}
table( colData(sce)$libraryID )
```

- Repeat to find one or two patterns you find interesting

```{r}

```

# Document software

```{r}
sessionInfo()
```

